Model for DES Type Ia Supernova Cosmology Analysis

### 2015-08-07 doc/model.tex
clean up code getting rid of commented stuff, fix documents to revised view of model

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
that have passed some selection criteria (e.g.\ classified as Type~Ia supernovae, SNe~Ia).  The probability distribution
function (pdf) of the Hubble diagram is denoted as
\begin{equation}
p({\mu},{z} |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S;
{\text{RA}}, {\text{Dec}}, \text{Detected}, {\text{Selected}}).
\label{hd:eqn}
\end{equation}
The semicolon separates random and fixed parameters.
The set of distance moduli
and redshifts $(\mu, z)$ is the Hubble diagram.  The available information
from which the Hubble diagram is inferred include:
the transient photometry ${ADU}$; spectral measurements of
transient
type, redshift and SN-parameters ${T}_S$, ${z}_S$, ${\theta}_S$;
mass, sSFR, etc.; transient coordinates  $\text{RA}$, $\text{Dec}$;
and the fact that the transients were Detected and Selected for the analysis.

The pdf in Equation~\ref{hd:eqn} is proportional to
\begin{equation}
p({\mu},{z}; \text{RA}, \text{Dec},  \text{Detected}, {\text{Selected}}
 |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S)
\label{hd2:eqn}
\end{equation}
and this is what we model.  A sketch of the model (with the optional inclusion
of calibration) is depicted in the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.
Note that although we have been discussing the model for the Hubble diagram, the same PGM can
 be used to express the pdf for cosmological
parameters, labelled as $\theta_\mu$.

@@

The model describes the transients that constitute the cosmology analysis, given the following
information.  
\begin{itemize}
\item ${\text{RA}}$, ${\text{Dec}}$.  The observed
coordinates.  For our purposes the observed, rather than true, coordinates are sufficient.
\item Detected.  The transient is discovered.
\item Selected. The transient is selected for the cosmology analysis.  This presumably
includes its classification as SN~Ia.
\end{itemize}

The universe is modeled as the background cosmology, transients that could
enter the sample, and their host galaxies.
\begin{itemize}
\item $P(g_i; {\text{RA}}/{\text{Dec}})$. The host galaxy $g_i$
is drawn based on coordinates.
It is drawn from
the union of the objects in the observed galaxy catalog
plus unobserved galaxies that have not
met the detection threshold, presumably due to their low surface brightness.
This pdf is
the deliverable we want from Ravi.
\item $z_i$, $\theta_{gi}$ are the galaxy redshift and parameters (e.g.\ mass, sSFR, metallicity).
For observed galaxies the parameter values are based on measurements e.g.\ $\theta_{gi} \sim \mathcal{N}({\theta}_{gi},\sigma^2_{{\theta}_{gi}})$, otherwise the distribution
of $\theta_{gi}$ for ``hostless'' galaxies is described by the model.
Redshifts may be photometric or spectroscopic.
\item $\theta_\mu$ are the parameters that specify that distance modulus.
\item $\mu_i=\mu(z_i; \theta_\mu)$.  The distance modulus is a function of redshift,
whose specific form is unspecified at the moment,  it could
be a $\Lambda$CDM prediction with cosmological parameters, a spline with knot values
as parameters.
The redshift of the galaxy is used in the Hubble diagram.
\item $P(T_i | \theta_{gi})$.  The transient type $T_i$, SN~Ia or non-Ia, depends
on the host properties.  We could make the relative rates be dependent
on specific star formation rate and morphology.
\item $p(\theta_T^{Ia}, \theta_{Ti}^{Ia})$.  The model parameters for SNe~Ia, including
parameters that describe distributions.
Different SN~Ia models could be considered: the treatment of SALT2 parameters
has been presented in \citet{2011MNRAS.418.2308M}.
The SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would constitute $\theta_{Ti}^{Ia}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would constitute $\theta_T^{Ia}$.
\item $p(\theta_T^{non-Ia}, \theta_{Ti}^{non-Ia})$.  Non-Ia transients
that can be mistyped as Ia are described by a parameterized model.
\item $L_i(t,\lambda; T_i, \theta_T^{Ia}, \theta_{Ti}^{Ia}, \theta_T^{non-Ia}, \theta_{Ti}^{non-Ia},
\theta_{gi})$.  The source model fixes
the luminosity. The  model includes  information on the
source and line-of-sight effects, which specify the effective SED.   Only the
parameters of the model of the appropriate type $T$ are relevant.  The galaxy parameters
$\theta_{gi}$ are included if correlations with host-galaxy properties are to be modeled.
\item $n_i(t,\lambda; L_i, \mu_i, z_i)$, $n_{gi}(t,\lambda; L_i, \mu_i, z_i)$.  The  fluxes of
the transient and host that are incident at Earth
are fixed by the luminosity, distance modulus, and redshift.
\end{itemize}

@@

The observations are as follows:
\begin{itemize}
\item ${\mathit{ADU}}$ is the realized flux can be expressed as
\begin{equation}
P({\mathit{ADU}} | \overline{\mathit{ADU}}, 
\overline{\mathit{ADU}}_g; \text{Dectected}, \text{Selected})=\frac{P({\mathit{ADU}};\text{Dectected}, \text{Selected} |  
\overline{\mathit{ADU}}, \overline{\mathit{ADU}}_g)}{P(\text{Dectected}, \text{Selected} | 
\overline{\mathit{ADU}},  \overline{{\mathit{ADU}}}_g)},
\end{equation}
which can be determined from detection and typing efficiency studies.
\item ${T}_S$, ${z}_S$, ${\theta}_S$ Transient type, redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
A very simple association is ${T}_S = T$, ${z}_S=z$, ${\theta}_S \in \theta_g$,
or a blurry version instead of a straight equality.
\item ${Z}$ represents the calibration measurements.   Although this may be given
by the Calibration Task Force, the PGM is written for the more general case where the calibration
is modeled within our analysis.
\end{itemize}

>>>>

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
that have passed some selection criteria (e.g.\ classified as Type~Ia supernovae, SNe~Ia).  The probability distribution
function (pdf) of the Hubble diagram is denoted as
\begin{equation}
p({\mu},{z} |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S, \theta_G,
{\text{Select}}).
\label{hd:eqn}
\end{equation}
The set of distance moduli
and redshifts $(\mu, z)$ is the Hubble diagram.  The available information
from which the Hubble diagram is inferred include:
the transient photometry ${ADU}$; spectral measurements of
transient
type, redshift and SN-parameters ${T}_S$, ${z}_S$, ${\theta}_S$;
host-galaxy parameter measurements $\theta_H$ including redshift,
mass, sSFR, etc.; and the sample-selection score Select.

A sketch of a more extensive model that includes the
model of Equation~\ref{hd:eqn} is depicted in the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.
Note that although we have been discussing the model for the Hubble diagram, the same PGM can
 be used to express the pdf for cosmological
parameters, labelled as $\theta_\mu$.

@@

The universe is modeled as the background cosmology, transients that could
enter the sample, and their host galaxies.
\begin{itemize}
\item  $g_i$ labels the distinct galaxy that is the transient host.
\item $z_i$, $\theta_{gi}$ are the galaxy redshift and parameters (e.g.\ mass, sSFR, metallicity).
Given the galaxy $g_i$, these values are specified.
\item $\theta_\mu$ are the parameters that specify that distance modulus.
\item $\mu_i=\mu(z_i; \theta_\mu)$.  The distance modulus is a function of redshift,
whose specific form is unspecified at the moment,  it could
be a $\Lambda$CDM prediction with cosmological parameters, a spline with knot values
as parameters.
The redshift of the galaxy is used in the Hubble diagram.
\item $P(T_i | \theta_{gi})$.  The transient type $T_i$, SN~Ia or non-Ia, depends
on the host properties.  We could make the relative rates be dependent
on specific star formation rate and morphology.
\item $p(\theta_T^{Ia}, \theta_{Ti}^{Ia})$.  The model parameters for SNe~Ia, including
parameters that describe distributions.
Different SN~Ia models could be considered: the treatment of SALT2 parameters
has been presented in \citet{2011MNRAS.418.2308M}.
The SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would constitute $\theta_{Ti}^{Ia}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would constitute $\theta_T^{Ia}$.
\item $p(\theta_T^{non-Ia}, \theta_{Ti}^{non-Ia})$.  Non-Ia transients
that can be mistyped as Ia are described by a parameterized model.
\item $L_i(t,\lambda; T_i, \theta_T^{Ia}, \theta_{Ti}^{Ia}, \theta_T^{non-Ia}, \theta_{Ti}^{non-Ia},
\theta_{gi})$.  The source model fixes
the luminosity. The  model includes  information on the
source and line-of-sight effects, which specify the effective SED.   Only the
parameters of the model of the appropriate type $T$ are relevant.  The galaxy parameters
$\theta_{gi}$ are included if correlations with host-galaxy properties are to be modeled.
\item $n_i(t,\lambda; L_i, \mu_i, z_i)$, $n_{gi}(t,\lambda; L_i, \mu_i, z_i)$.  The  fluxes of
the transient and host that are incident at Earth
are fixed by the luminosity, distance modulus, and redshift.
\end{itemize}

@@

The observations are as follows:
\begin{itemize}
\item ${\mathit{ADU}}$ is the realized flux. 
\item Select is the score used to decide if the transient is used in the analysis.
It and $\mathit{ADU}$ are not independent so what is relevant is their joint distribution
\begin{equation}
P({\mathit{ADU},\text{Select}} | \overline{\mathit{ADU}}, 
\overline{\mathit{ADU}}_g),
\end{equation}
which can be determined from detection and typing efficiency studies.
\item $\theta_G$ represents the galaxy parameters of the
galaxy catalog used to identify prospective host galaxies.
The contribution to the likelihood $P(\theta_G|g_i)$ can be
estimated as $P(g_i|\theta_G)\frac{P(\theta_G)}{P(z)}$: Ravi is determining $P(g_i|\theta_G)$.
\item ${T}_S$, ${z}_S$, ${\theta}_S$ Transient type, redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
A very simple association is ${T}_S = T$, ${z}_S=z$, ${\theta}_S \in \theta_g$,
or a blurry version instead of a straight equality.
\item ${Z}$ represents the calibration measurements.   Although this may be given
by the Calibration Task Force, the PGM is written for the more general case where the calibration
is modeled within our analysis.
\end{itemize}
